Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-08-08 doc/JAIR/jair-paper.tex
Updates

One approach to model counting, known as \emph{knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the \emph{deterministic, decomposable, negation normal form}
(det-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004}, as well as the more restricted
\emph{decision-DNNF} (dec-DNNF)~\cite{huang:jair:2007,oztok:cp:2014},
represent a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And or Or operation.  Restrictions
are placed on the structure of the graph (described in Section~\ref{sect:pog}) such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a det-DNNF representation.  These include standard and weighted model
counting, and much more.

@@

\begin{dfn}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated \emph{dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of \emph{models} $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its \emph{negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{dfn}

@@

\begin{dfn}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
  A \emph{partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to each product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to each sum must have disjoint models.  That is, operation
        $\bigvee_{1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
\end{dfn}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

\begin{dfn}[Commutative Ring]
  A \emph{commutative ring} $\ring$ is an algebraic structure
  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$,
  with elements in the set $\dset$ and with commutative and
  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
  such that multiplication distributes
  over addition.  $\addident$ is the additive identity and $\mulident$ is
  the multiplicative identity.  Every element $a \in \dset$ has an
  \emph{additive inverse} $-a$ such that $a + -a = \addident$.
\label{def:ring}
\end{dfn}
We write $a - b$ as a shorthand for $a + -b$.

@@

We describe the structural approach to proof generation as a recursive procedure
$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
node $\nodeu$, partial assignment
$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
The procedure adds a number of clauses to the proof, culminating with
the addition of the \emph{target} clause:
$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
that $(\bigwedge_{\lit \in \passign} \lit) \imply u$, i.e.,
that any total
assignment $\assign$ such that $\passign \subseteq \assign$
will assign $\assign(u) = 1$.
The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
The result will therefore be to add unit clause $\{r\}$ to the proof.
Here we present a correct, but somewhat inefficient formulation of
$\validate$.  We then refine it with some optimizations.

>>>>

One approach to model counting, known as \emph{knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the \emph{deterministic, decomposable, negation normal form}
(det-DNNF) introduced by
\citeA{darwiche:aaai:2002,darwiche:ecai:2004}, as well as the more restricted
\emph{decision-DNNF} (dec-DNNF)~\cite{huang:jair:2007,oztok:cp:2014},
represent a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And or Or operation.  Restrictions
are placed on the structure of the graph (described in Section~\ref{sect:pog}) such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a det-DNNF representation.  These include standard and weighted model
counting, and much more.

@@

\begin{dfn}
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated \emph{dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of \emph{models} $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its \emph{negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{dfn}

@@

\begin{dfn}
  \label{def:partitioned-operation-formula}
  A \emph{partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to each product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to each sum must have disjoint models.  That is, operation
        $\bigvee_{1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
\end{dfn}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

\begin{dfn}
  A \emph{commutative ring} $\ring$ is an algebraic structure
  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$,
  with elements in the set $\dset$ and with commutative and
  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
  such that multiplication distributes
  over addition.  $\addident$ is the additive identity and $\mulident$ is
  the multiplicative identity.  Every element $a \in \dset$ has an
  \emph{additive inverse} $-a$ such that $a + -a = \addident$.
\label{def:ring}
\end{dfn}
We write $a - b$ as a shorthand for $a + -b$.

@@

The structural approach to proof generation takes the form of a recursive procedure
$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
node $\nodeu$, partial assignment
$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
The procedure adds a number of clauses to the proof, culminating with
the addition of the \emph{target} clause:
$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
that $(\bigwedge_{\lit \in \passign} \lit) \imply u$, i.e.,
that any total
assignment $\assign$ such that $\passign \subseteq \assign$
will assign $\assign(u) = 1$.
The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
The result will therefore be to add unit clause $\{r\}$ to the proof.
Here we present a correct, but somewhat inefficient formulation of
$\validate$.  We then refine it with some optimizations.
